Getting mortality from atlantisom: testing calc_Z function II

### 2019-06-07 TestCalcZ2.Rmd
initial comparisons calc Z with mort.txt

---
title: "Getting mortality from atlantisom: testing calc_Z function I"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
bibliography: "packages.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'knitr', 'rmarkdown', 'tidyr', 'dplyr', 'ggplot2',
  'data.table', 'here', 'ggforce', 'ggthemes', 'r4ss'
), 'packages.bib')
```

@@

This page documents initial testing of the atlantisom package in development at https://github.com/r4atlantis/atlantisom using three different [Atlantis](https://research.csiro.au/atlantis/) output datasets. Development of atlantisom began at the [2015 Atlantis Summit](https://research.csiro.au/atlantis/atlantis-summit/) in Honolulu, Hawaii, USA. 
On this page we demonstrate use of atlantisom on Norwegian Barents Sea Atlantis (NOBA) output files to test function `calc_Z` calculating total mortality. This is used both to derive "natural mortality" for comparison with stock assessments and to split age classes into true ages using the function `calc_stage2age`.

@@

@@

initCCA <- FALSE
initNEUS <- FALSE
initNOBA <- TRUE

@@

if(initCCA) source(here("config/CC1Config.R"))

@@

```{r loadcatchb}

@@

@@

This section tests the `atlantisom::calc_Z()` function step by step. Ultimately this output goes to `calc_stage2age` to produce annual numbers at true age, which we can compare with the ANNAGEBIO.nc output produced by NOBA.

@@

surveyN_ss <- surveyN[surveyN$species == "North_atl_cod",]

plotN <-ggplot() +
  geom_line(data=surveyN_ss, aes(x=time/5,y=atoutput, color="survey census N"), 
            alpha = 10/10) +
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

```{r truenatagecl, eval=FALSE}
# We need an effective N for the sample fish function, setting equal to actual N
# this one is high but not equal to total for numerous groups
effNhigh <- data.frame(species=survspp, effN=rep(1e+8, length(survspp)))

```{r Natage-plot}

@@

]The atlantisom function `calc_stage2age` calls the `calc_Z` function, applying a total mortality rate to estimate the numbers at true age within an age class, but it is unclear whether to use this on the aggregated numbers or the full resolution numbers at age. Here we test outputs of Z for several levels of aggregation:

@@

# cut to a single species in YOY file, hardcoded for NOBA cod
YOY_ss <- YOY %>%
  select(Time, NCO.0)

# numbers at agecl at full resolution (all polygons and layers)
truenums_ss <- truth$nums[truth$nums$species == "North_atl_cod",]

@@

# numbers at ageclaggregated over layers, retain polygons (output of create_survey)
survey_testN_ss <- survey_testNall[survey_testNall$species == "North_atl_cod",]

# numbers at agecl aggregated over layers and polygons (output of sample_fish)
# done above
# Natage_ss <- Natage[Natage$species == "North_atl_cod",]

@@

Fixed a bug where grep was used if a # appeared, none of those in NOBA, will have to see if output is produced in CCA. This ran for the survey resolution numbers. I then went back and ran the two other levels above after fixing `calc_Z.`

Something is wrong with running after `sample_fish` because sampleresZ returned all 0s. Both true full resolution numbers and survey resolution numbers (aggregated over polygon) returned nonzero Z values. They are identical aside from seven values with no apparent pattern (rows 84, 85, 130, 131, 147, 172, 196) out of 201 that differ at e-16. I would be happer if the differences were all exactly 0, but I suppose we can attribute that to rounding error.

@@

### What is F at each timestep?
This probably depends on how the model is set up, but for models forced with F by species we ought to be able to read this out. If it is a daily rate then we mulitply by toutinc to get the rate at each of these output timestep Z values.

### Quick test to see if Z is right?
Perhaps overly optimistic, but if we feed these into `calc_stage2age` and get a result that matches the NOBA output ANNAGEBIO.nc, then we can assume the Zs are correct.

@@

```{r plotannage}

@@

```{r compareages}

@@

```{r plot-esttrueage-atltrueage}

>>>>

---
title: "Getting mortality from atlantisom: testing calc_Z function II"
author: "Sarah Gaichas and Christine Stawitz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

This page documents initial testing of the atlantisom package in development at https://github.com/r4atlantis/atlantisom using three different [Atlantis](https://research.csiro.au/atlantis/) output datasets. Development of atlantisom began at the [2015 Atlantis Summit](https://research.csiro.au/atlantis/atlantis-summit/) in Honolulu, Hawaii, USA. 
On this page we demonstrate use of atlantisom on both California Current (CCA) and Norwegian Barents Sea Atlantis (NOBA) output files to test function `calc_Z` calculating total mortality. This is used both to derive "natural mortality" for comparison with stock assessments and to split age classes into true ages using the function `calc_stage2age`.

@@

We can also compare both models with the output[scenario.name]Mort.txt file, which reports annual M and F. The user's manual notes: 
>This file is currently only useful for looking at relative M vs F values for a species, as itdoes not give accurate mortalities

So we can assume that the scale of Z from this file will not be the same as our calculation, but it should be useful for splitting out F to get M, and for comparing overall patterns.

We will first use CCA here because the contrasting F scenario has been implemented. We will test with both sardines (M needed, but not age splitting) and hake (M and age splitting needed).

@@

initCCA <- TRUE
initNEUS <- FALSE
initNOBA <- FALSE

species_ss <- c("Pacific_sardine")

@@

if(initCCA) source(here("config/CC2Config.R"))

@@

<!-- not sure we need catch bio for this
```{r loadcatchb}

@@

truecatchbio_ss <- truecatchbio[truecatchbio$species == species_ss,]
# note: time output of this file is in days
# what model timestep is this output matching?
# rescale catch in biomass time from days to years
truecatchbio_ss$time <- as.integer(truecatchbio_ss$time/365)

```
-->

@@

This section tests the `atlantisom::calc_Z()` function step by step. Ultimately this output goes to `calc_stage2age` to produce annual numbers at true age, which we can compare with the ANNAGEBIO.nc output produced by NOBA. For now we will just compare to the Mort.txt output.

@@

surveyN_ss <- surveyN[surveyN$species == species_ss,]

plotN <-ggplot() +
  geom_line(data=surveyN_ss, aes(x=time/stepperyr,y=atoutput, color="survey census N"), 
            alpha = 10/10) +
  theme_tufte() +
  theme(legend.position = "top") +
  labs(colour=scenario.name)

@@

@@

The atlantisom function `calc_stage2age` calls the `calc_Z` function, applying a total mortality rate to estimate the numbers at true age within an age class, but it is unclear whether to use this on the aggregated numbers or the full resolution numbers at age. Here we test outputs of Z for several levels of aggregation:

@@

# get code matching species name to split YOY file
code_ss <- funct.groups$Code[which(funct.groups$Name == species_ss)]

# cut to a single species in YOY file
YOY_ss <- YOY %>%
  select(Time, paste0(code_ss, ".0"))

# numbers at agecl at full resolution (all polygons and layers)
truenums_ss <- truth$nums[truth$nums$species == species_ss,]

@@

# numbers at agecl aggregated over layers, retain polygons (output of create_survey)
survey_testN_ss <- survey_testNall[survey_testNall$species == species_ss,]

# numbers at agecl aggregated over layers and polygons
Natage_ss <- readRDS(file.path(d.name, paste0(scenario.name, "natage_census_sard.rds")))

@@

Fixed a bug where grep was used if a # appeared. This ran for the survey resolution numbers. I then went back and ran the two other levels above after fixing `calc_Z.`

Both true full resolution numbers and survey resolution numbers (aggregated over polygon) returned nonzero Z values. As with previous the NOBA test, they are identical aside from eight values with no apparent pattern (rows 100, 105, 180, 200, 220, 240, 355, 400) out of 500 that differ at e-16. Same story with difference from sample resolution; this does not matter to output.

```{r differentZ}
fullresZ$atoutput - surveyresZ$atoutput

fullresZ$atoutput - sampleresZ$atoutput
```

I would be happer if the differences were all exactly 0, but I suppose we can attribute that to rounding error.

### What is F at each timestep?
A relative annual F is found in outputMort.txt!
```{r getrelF}

file.mort <- file.path(d.name, paste0("output", scenario.name, "Mort.txt"))

mortish <- read.table(file.mort, header = TRUE)

relF_ss <- mortish %>%
  select(Time, relF = paste0(code_ss, ".F"))

relM_ss <- mortish %>%
  select(Time, relM = paste0(code_ss, ".M"))

plotF <- ggplot(data=relF_ss, aes(x=Time/365, y=relF)) +
  geom_line() +
  theme_tufte() +
  labs(subtitle = paste(scenario.name, species_ss))

plotF

This is consistent with the scenario implemented. So far so good.

This probably depends on how the model is set up, but for models forced with F by species we ought to be able to read this out from an input file. If it is a daily rate then we mulitply by toutinc to get the rate at each of these output timestep Z values.

### Quick test to see if Z is right?
Compare with outputMort.txt!
```{r compZ}

Zish <- merge(relF_ss,relM_ss) %>%
  mutate(relZ = relF + relM)

plotZ <- ggplot() +
  geom_line(data=Zish, aes(x=Time/365, y=relZ)) +
  geom_point(data=fullresZ, aes(x=(time*73)/365, y=atoutput)) +
  theme_tufte() +
  labs(subtitle = paste(scenario.name, species_ss))

plotZ
```

Nope. We have a constant value punctuated by occassional variation. Seems wrong. Variation doesn't look anything like the txt file either. Perhaps the 5x yr timesteps are messing this up?

### How to turn timestep Z into annual Z for comparison?

Time to step through line by line again.

@@

Might be a good time to switch back to NOBA for comparison with true annual ages.

Everything below here is old.

```{r truenatagecl, eval=FALSE}
# We need an effective N for the sample fish function, setting equal to actual N
# this one is high but not equal to total for numerous groups

# this doesnt work for sardines in CCA, change to full census

effNhigh <- data.frame(species=survspp, effN=rep(1e+8, length(survspp)))

```{r Natage-plot, eval=FALSE}

Perhaps overly optimistic, but if we feed these into `calc_stage2age` and get a result that matches the NOBA output ANNAGEBIO.nc, then we can assume the Zs are correct.

@@

```{r plotannage, eval=FALSE}

@@

```{r compareages, eval=FALSE}

@@

```{r plot-esttrueage-atltrueage, eval=FALSE}
